Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-05-13 article-models-ttderived.tex
a few more tags added
commenting stuff so that the article becomes shorter
nearly finished with the section about fiber inclusions

To find out what the
apropriate way is, note that
\begin{align*}
\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}
& \jdeq
  \ctxext{\ctxwk{\ctxext{A}{P}}{A}}{\ctxwk{\ctxext{A}{P}}{P}}
  \tag{by \autoref{comp-we-f}}
  \\
& \jdeq
  \ctxext{\ctxwk{P}{{A}{A}}}{\ctxwk{P}{{A}{P}}}
  \tag{by \autoref{comp-ew-f}}
\end{align*}
We have the term $\jterm{{\Gamma}{{A}{P}}}{\ctxwk{P}{A}}{\ctxwk{P}{\idtm{A}}}$.
Thus we need to find out what $\subst{\ctxwk{P}{\idtm{A}}}{\ctxwk{P}{{A}{P}}}$ is:
\begin{align*}
\subst{\ctxwk{P}{\idtm{A}}}{\ctxwk{P}{{A}{P}}} 
& \jdeq 
  \ctxwk{P}{\subst{\idtm{A}}{\ctxwk{A}{P}}}
  \\
& \jdeq 
  \ctxwk{P}{P},
\end{align*}
where we find the term $\idtm{P}$. Therefore we define:

@@

\begin{proof}
To prove the first judgmental equality, note that
\begin{align*}
w 
& \jdeq 
  \subst{w}{\idtm{\ctxext{A}{P}}} 
  \tag{by \autoref{idfunc-subst-defn}}\\
& \jdeq 
  \subst
    { w}
    { { \idtm{P}}
      { { \ctxwk{P}{\idtm{A}}}
        { \idtm{\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}}}
        }
      }
  \tag{by \autoref{idfunc-ext-comp}}\\
& \jdeq 
  \subst
    { {w}
      {\idtm{P}}
      }
    { {w}
      { { \ctxwk{P}{\idtm{A}}
          }
        { \idtm{\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}}
          }
        }
      }
  \tag{by \autoref{comp-ss-t}}\\
& \jdeq 
  \subst
    { {w}
      {\idtm{P}}
      }
    { { {w}
        {\ctxwk{P}{\idtm{A}}}
        }
      { {w}
        {\idtm{\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}}}
        }
      }
  \tag{by \autoref{comp-ss-t}}\\
& \jdeq 
  \subst
    { {w}
      {\idtm{P}}
      }
    { { {w}
        {\ctxwk{P}{\idtm{A}}}
        }
      { {w}
        {\ctxwk{\ctxext{A}{P}}{\idtm{\ctxext{A}{P}}}}
        }
      }
  \tag{by \autoref{comp-wi-t}}\\
& \jdeq 
  \subst
    { {w}
      {\idtm{P}}
      }
    { { {w}
        {\ctxwk{P}{\idtm{A}}}
        }
      { \idtm{\ctxext{A}{P}}
        }
      }
  \tag{by \autoref{cancellation-ws-t}}\\
& \jdeq 
  \tmext{A}{P}{\cprojfst{A}{P}{w}}{\cprojsnd{A}{P}{w}}
  \tag{by definition}
\end{align*}
To prove the second judgmental equality, let $\jterm{\Gamma}{A}{x}$ and
$\jterm{\Gamma}{\subst{x}{P}}{u}$. Then we have
\begin{align*}
\cprojfst{A}{P}{\tmext{A}{P}{x}{u}}
& \jdeq 
  \subst{\tmext{A}{P}{x}{u}}{\ctxwk{P}{\idtm{A}}}
  \\
& \jdeq 
  \subst{u}{{x}{\ctxwk{P}{\idtm{A}}}} 
  \\
& \jdeq 
  \subst{u}{\ctxwk{\subst{x}{P}}{\subst{x}{\idtm{A}}}}
  \\
& \jdeq 
  \subst{x}{\idtm{A}}
  \\
& \jdeq 
  x.
\end{align*}
To prove the third judgmental equality, note that
\begin{align*}
\cprojsnd{A}{P}{\tmext{A}{P}{x}{u}}
& \jdeq 
  \subst{\tmext{A}{P}{x}{u}}{\idtm{P}}
  \\
& \jdeq 
  \subst{u}{{x}{\idtm{P}}}
  \\
& \jdeq 
  \subst{u}{\idtm{\subst{x}{P}}}
  \\
& \jdeq 
  u.
  \qedhere
\end{align*}
\end{proof}

@@

@@

@@

@@

\subsection{Another special case of projections}
In this subsection we investigate the special case of a projection which
appears as a morphism from $\ctxext{{A}{P}}{\ctxwk{P}{Q}}$ to $\ctxext{A}{Q}$
in context $\Gamma$, where we assume to have the families 
$\jfam{{\Gamma}{A}}{P}$ and $\jfam{{\Gamma}{A}}{Q}$.

@@

\begin{lem}
We have the judgmental equality
\begin{equation*}
\jfameq
  {{{{\Gamma}{A}}{P}}{\ctxwk{P}{\mfam{A}}}}
  {\jcomp{}{\bar{\typefont{pr}}}{Q}}
  {\ctxwk{P}{Q}}
\end{equation*}
for any family $Q$ of contexts over $\ctxext{{\Gamma}{A}}{\mfam{A}}$ 
\end{lem}

>>>>

To find out what the
apropriate way is, note that
\begin{align*}
\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}
& \jdeq
  \ctxext{\ctxwk{\ctxext{A}{P}}{A}}{\ctxwk{\ctxext{A}{P}}{P}}
  \tag{by \autoref{comp-we-f}}
  \\
& \jdeq
  \ctxext{\ctxwk{P}{{A}{A}}}{\ctxwk{P}{{A}{P}}}
  \tag{by \autoref{comp-ew-f}}
\end{align*}
We have the term $\jterm{{\Gamma}{{A}{P}}}{\ctxwk{P}{A}}{\ctxwk{P}{\idtm{A}}}$.
Thus we need to find out what $\subst{\ctxwk{P}{\idtm{A}}}{\ctxwk{P}{{A}{P}}}$ is:
\begin{align*}
\subst{\ctxwk{P}{\idtm{A}}}{\ctxwk{P}{{A}{P}}} 
& \jdeq 
  \ctxwk{P}{\subst{\idtm{A}}{\ctxwk{A}{P}}}
  \tag{by \autoref{comp-ws-f}}
  \\
& \jdeq 
  \ctxwk{P}{P},
  \tag{by \autoref{idfunc-wk-defn}}
\end{align*}
where we find the term $\idtm{P}$. Therefore we define:

@@

\begin{proof}
To prove the first judgmental equality, note that
\begin{align*}
w 
& \jdeq 
  \subst{w}{\idtm{\ctxext{A}{P}}} 
  \tag{by \autoref{cancellation-si}}\\
& \jdeq 
  \subst
    { w}
    { { \idtm{P}}
      { { \ctxwk{P}{\idtm{A}}}
        { \idtm{\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}}}
        }
      }
  \tag{by \autoref{idfunc-ext-comp}}\\
& \jdeq 
  \subst
    { {w}
      {\idtm{P}}
      }
    { {w}
      { { \ctxwk{P}{\idtm{A}}
          }
        { \idtm{\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}}
          }
        }
      }
  \tag{by \autoref{comp-ss-t}}\\
& \jdeq 
  \subst
    { {w}
      {\idtm{P}}
      }
    { { {w}
        {\ctxwk{P}{\idtm{A}}}
        }
      { {w}
        {\idtm{\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}}}
        }
      }
  \tag{by \autoref{comp-ss-t}}\\
& \jdeq 
  \subst
    { {w}
      {\idtm{P}}
      }
    { { {w}
        {\ctxwk{P}{\idtm{A}}}
        }
      { {w}
        {\ctxwk{\ctxext{A}{P}}{\idtm{\ctxext{A}{P}}}}
        }
      }
  \tag{by \autoref{comp-wi-t}}\\
& \jdeq 
  \subst
    { {w}
      {\idtm{P}}
      }
    { { {w}
        {\ctxwk{P}{\idtm{A}}}
        }
      { \idtm{\ctxext{A}{P}}
        }
      }
  \tag{by \autoref{cancellation-ws-t}}\\
& \jdeq 
  \tmext{A}{P}{\cprojfst{A}{P}{w}}{\cprojsnd{A}{P}{w}}
  \tag{by definition}
\end{align*}
To prove the second judgmental equality, let $\jterm{\Gamma}{A}{x}$ and
$\jterm{\Gamma}{\subst{x}{P}}{u}$. Then we have
\begin{align*}
\cprojfst{A}{P}{\tmext{A}{P}{x}{u}}
& \jdeq 
  \subst{\tmext{A}{P}{x}{u}}{\ctxwk{P}{\idtm{A}}}
  \tag{by definition}
  \\
& \jdeq 
  \subst{u}{{x}{\ctxwk{P}{\idtm{A}}}} 
  \tag{by \autoref{comp-es}}
  \\
& \jdeq 
  \subst{u}{\ctxwk{\subst{x}{P}}{\subst{x}{\idtm{A}}}}
  \tag{by \autoref{comp-sw-t}}
  \\
& \jdeq 
  \subst{x}{\idtm{A}}
  \tag{by \autoref{cancellation-ws-t}}
  \\
& \jdeq 
  x.
  \tag{by \autoref{cancellation-si}}
\end{align*}
To prove the third judgmental equality, note that
\begin{align*}
\cprojsnd{A}{P}{\tmext{A}{P}{x}{u}}
& \jdeq 
  \subst{\tmext{A}{P}{x}{u}}{\idtm{P}}
  \tag{by definition}
  \\
& \jdeq 
  \subst{u}{{x}{\idtm{P}}}
  \tag{by \autoref{comp-es}}
  \\
& \jdeq 
  \subst{u}{\idtm{\subst{x}{P}}}
  \tag{by \autoref{comp-si-t}}
  \\
& \jdeq 
  u.
  \tag{by \autoref{cancellation-si}}
\end{align*}
\end{proof}

@@

We can immediately show that composing with a fiber inclusion is substitution.
Thus, the following lemma asserts that we have a family pullback square
\begin{equation*}
\begin{tikzcd}
\subst{x}{Q} \ar[fib]{d} \ar{r} & Q \ar[fib]{d} \\
\subst{x}{P} \ar{r}[swap]{\finc{x}{P}} & \ctxext{A}{P}
\end{tikzcd}
\end{equation*}
in context $\Gamma$.
The upper arrow is just the identity morphism over $\finc{x}{P}$.

\begin{lem}\label{lem:finc-precomp}
Composition with $\finc{x}{P}$ is (the action on families of)
substitution with $x$:
\begin{align*}
& \inference
  { \jterm{\Gamma}{A}{x}
    \jfam{{{\Gamma}{A}}{P}}{Q}
    }
  { \jfameq
      {{\Gamma}{\subst{x}{P}}}
      {\jcomp{\subst{x}{P}}{\finc{x}{P}}{Q}}
      {\subst{x}{Q}}
    }
  \\
& \inference
  { \jterm{\Gamma}{A}{x}
    \jfam{{{{\Gamma}{A}}{P}}{Q}}{R}
    }
  { \jfameq
      {{{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}}
      {\jcomp{\subst{x}{P}}{\finc{x}{P}}{R}}
      {\subst{x}{R}}
    }
  \\
& \inference
  { \jterm{\Gamma}{A}{x}
    \jterm{{{{\Gamma}{A}}{P}}{Q}}{R}{h}
    }
  { \jtermeq
      {{{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}}
      {\subst{x}{R}}
      {\jcomp{\subst{x}{P}}{\finc{x}{P}}{h}}
      {\subst{x}{h}}
    }
\end{align*}
\end{lem}

\begin{proof}
We only prove the validity of the first inference rule. Let $\jterm{\Gamma}{A}{x}$
be a term and let $\jfam{{{\Gamma}{A}}{P}}{Q}$ be a family. Then we have the
judgmental equalities
\begin{align*}
\jcomp{\subst{x}{P}}{\finc{x}{P}}{Q}
& \jdeq
  \unfold{\jcomp{\subst{x}{P}}{\unfold{\finc{x}{P}}}{Q}}
  \tag{by definition}
  \\
& \jdeq
  \subst
    {\idtm{\subst{x}{P}}}
    {{\ctxwk{\subst{x}{P}}{x}}{\ctxwk{\subst{x}{P}}{Q}}}
  \tag{by \autoref{comp-es}}
  \\
& \jdeq
  \subst
    {\idtm{\subst{x}{P}}}
    {\ctxwk{\subst{x}{P}}{\subst{x}{Q}}}
  \tag{by \autoref{comp-ws-f}}
  \\
& \jdeq
  \subst{x}{Q}.
  \tag{by \autoref{idfunc-precomp}}
\end{align*}
\end{proof}

We can give a second characterization of fiber inclusions:

\begin{lem}\label{lem:finc-char2}
We have the following inference rule
\begin{equation*}
\inference
  { \jterm{\Gamma}{A}{x}
    \jfam{{\Gamma}{A}}{P}
    }
  { \jhomeq{\Gamma}{\subst{x}{P}}{{A}{P}}{\finc{x}{P}}{\subst{x}{\idtm{{A}{P}}}}
    } 
\end{equation*}
\end{lem}

\begin{proof}
The proof is the following calculation:
\begin{align*}
\finc{x}{P}
& \jdeq 
  \tmext{\ctxwk{\subst{x}{P}}{x}}{\idtm{\subst{x}{P}}}
  \tag{by definition}
  \\
& \jdeq
  \subst
    { \idtm{\subst{x}{P}}
      }
    { { \ctxwk{\subst{x}{P}}{x}
        }
      { \idtm{\ctxext{\ctxwk{\subst{x}{P}}{A}}{\ctxwk{\subst{x}{P}}{P}}}
        }
      }
  \tag{by definition}
  \\
& \jdeq
  \subst
    { \idtm{\subst{x}{P}}
      }
    { { \ctxwk{\subst{x}{P}}{x}
        }
      { \idtm{\ctxwk{\subst{x}{P}}{\ctxext{A}{P}}}
        }
      }
  \tag{by \autoref{comp-we-f}}
  \\
& \jdeq
  \subst
    { \idtm{\subst{x}{P}}
      }
    { { \ctxwk{\subst{x}{P}}{x}
        }
      { \ctxwk{\subst{x}{P}}{\idtm{\ctxext{A}{P}}}
        }
      }
  \tag{by \autoref{comp-wi-t}}
  \\
& \jdeq
  \subst
    { \idtm{\subst{x}{P}}
      }
    { \ctxwk
        { \subst{x}{P}
          }
        { \subst{x}{\idtm{\ctxext{A}{P}}}
          }
      }
  \tag{by \autoref{comp-ws-t}}
  \\
& \jdeq
  \subst{x}{\idtm{\ctxext{A}{P}}}.
  \tag{by \autoref{idfunc-precomp}}
\end{align*}
\end{proof}

\begin{cor}
The following inference rule is valid:
\begin{equation*}
\inference
  { \jterm{\Gamma}{A}{x}
    \jterm{\Gamma}{\subst{x}{P}}{u}
    }
  { \jtermeq{\Gamma}{{A}{P}}{\subst{u}{\finc{x}{P}}}{\tmext{x}{u}}
    }
\end{equation*}
\end{cor}

@@

\begin{proof}
Let $\jterm{\Gamma}{A}{x}$. We have the judgmental equalities
\begin{align*}
\finc{x}{\emptyf}
& \jdeq
  \unfold{\finc{x}{\emptyf}}
  \tag{by definition}
  \\
& \jdeq
  \tmext{\ctxwk{\emptyf}{x}}{\idtm{\emptyf}}
  \tag{by \autoref{comp-s0-c}}
  \\
& \jdeq
  \tmext{x}{\idtm{\emptyf}}
  \tag{by \autoref{comp-0w-t}}
  \\
& \jdeq
  \tmext{x}{\emptytm}
  \tag{by \autoref{comp-00-t}}
  \\
& \jdeq
  x
  \tag{by \autoref{lem:tmext-emp}}
\end{align*}
to prove the first judgmental equality. For the second, we have
\begin{align*}
\finc{\emptytm}{A}
& \jdeq
  \unfold{\finc{\emptytm}{A}}
  \tag{by definition}
  \\
& \jdeq
  \tmext{\emptytm}{\idtm{\subst{\emptytm}{A}}}
  \tag{by \autoref{comp-00-t}}
  \\
& \jdeq
  \idtm{\subst{\emptytm}{A}}
  \tag{by \autoref{lem:tmext-emp}}
  \\
& \jdeq
  \idtm{A}
  \tag{by \autoref{comp-0s-f}}.
\end{align*}
\end{proof}

@@

\begin{proof}
Let $\jterm{\Gamma}{A}{x}$, $\jterm{\Gamma}{\subst{x}{P}}{u}$ and
$\jfam{{{\Gamma}{A}}{P}}{Q}$. Then we have the judgmental equalities
\begin{align*}
\finc{\tmext{x}{u}}{Q}
& \jdeq
  \subst{\tmext{}{}{x}{u}}{\idtm{{{A}{P}}{Q}}}
  \tag{by \autoref{lem:finc-char2}}
  \\
& \jdeq
  \subst{u}{{x}{\idtm{{{A}{P}}{Q}}}}
  \tag{by \autoref{comp-es}}
  \\
& \jdeq
  \subst{u}{{x}{\idtm{{A}{{P}{Q}}}}}
  \tag{by \autoref{comp-ee-c}}
  \\
& \jdeq
  \subst{u}{\finc{x}{\ctxext{P}{Q}}}
  \tag{by \autoref{lem:finc-char2}}
  \\
& \jdeq
  \jcomp{}{\finc{u}{\subst{x}{Q}}}{\finc{x}{\ctxext{P}{Q}}}
  \tag{by \autoref{lem:finc-precomp}}
\end{align*}
\end{proof}

@@

\begin{comment}
\subsection{Another special case of projections}
In this subsection we investigate the special case of a projection which
appears as a morphism from $\ctxext{{A}{P}}{\ctxwk{P}{Q}}$ to $\ctxext{A}{Q}$
in context $\Gamma$, where we assume to have the families 
$\jfam{{\Gamma}{A}}{P}$ and $\jfam{{\Gamma}{A}}{Q}$.

@@

\begin{lem}
We have the judgmental equality
\begin{equation*}
\jfameq
  {{{{\Gamma}{A}}{P}}{\ctxwk{P}{\mfam{A}}}}
  {\jcomp{}{\bar{\typefont{pr}}}{Q}}
  {\ctxwk{P}{Q}}
\end{equation*}
for any family $Q$ of contexts over $\ctxext{{\Gamma}{A}}{\mfam{A}}$ 
\end{lem}
\end{comment}
